---
title: "Forest Fires Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Forest Fires Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/shubhammeshram579/forest-fires-prediction\>*

### Reference:

*\<Meshram, S. (n.d.). Forest Fires Prediction [Data set]. Kaggle. https://www.kaggle.com/datasets/shubhammeshram579/forest-fires-prediction\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Loading Dataset}
# Load dataset
forest_fire_data <- read.csv("forest_fires.csv", colClasses = c(
  X = "numeric",
  Y = "numeric",
  month = "factor",
  day = "factor",
  FFMC = "numeric",
  DMC = "numeric",
  DC = "numeric",
  ISI = "numeric",
  temp = "numeric",
  RH = "numeric",
  wind = "numeric",
  rain = "numeric",
  area = "numeric"
))

# Display the structure of the dataset
str(forest_fire_data)

# View the first few rows of the dataset
head(forest_fire_data)

# View the dataset in a separate viewer window
View(forest_fire_data)
```

## Measures of Frequency
```{r MOF}
# Measures of Frequency
# Frequency of occurrences for month
month_freq <- table(forest_fire_data$month)
print("Frequency of occurrences for month:")
print(month_freq)

# Frequency of occurrences for day
day_freq <- table(forest_fire_data$day)
print("Frequency of occurrences for day:")
print(day_freq)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean, Median, and Mode for temperature
temp_mean <- mean(forest_fire_data$temp)
temp_median <- median(forest_fire_data$temp)
temp_mode <- names(sort(table(forest_fire_data$temp), decreasing = TRUE))[1]
print("Measures of Central Tendency for temperature:")
print(paste("Mean:", temp_mean))
print(paste("Median:", temp_median))
print(paste("Mode:", temp_mode))

# Mean, Median, and Mode for humidity
RH_mean <- mean(forest_fire_data$RH)
RH_median <- median(forest_fire_data$RH)
RH_mode <- names(sort(table(forest_fire_data$RH), decreasing = TRUE))[1]
print("Measures of Central Tendency for humidity:")
print(paste("Mean:", RH_mean))
print(paste("Median:", RH_median))
print(paste("Mode:", RH_mode))
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Range, Variance, and Standard Deviation for temperature
temp_range <- range(forest_fire_data$temp)
temp_variance <- var(forest_fire_data$temp)
temp_sd <- sd(forest_fire_data$temp)
print("Measures of Distribution for temperature:")
print(paste("Range:", temp_range[2] - temp_range[1]))
print(paste("Variance:", temp_variance))
print(paste("Standard Deviation:", temp_sd))

# Range, Variance, and Standard Deviation for humidity
RH_range <- range(forest_fire_data$RH)
RH_variance <- var(forest_fire_data$RH)
RH_sd <- sd(forest_fire_data$RH)
print("Measures of Distribution for humidity:")
print(paste("Range:", RH_range[2] - RH_range[1]))
print(paste("Variance:", RH_variance))
print(paste("Standard Deviation:", RH_sd))

```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between temperature and humidity
temp_RH_correlation <- cor(forest_fire_data$temp, forest_fire_data$RH)
print("Correlation between temperature and humidity:")
print(temp_RH_correlation)

# Correlation between temperature and area
temp_area_correlation <- cor(forest_fire_data$temp, forest_fire_data$area)
print("Correlation between temperature and area:")
print(temp_area_correlation)
```

## ANOVA
```{r ANOVA}
# Perform ANOVA on temperature across different months
anova_result <- aov(temp ~ month, data = forest_fire_data)

# Print ANOVA summary
print(summary(anova_result))
```

## Plots
```{r Plots}
# Univariate Plots
# Histogram of temperature
hist(forest_fire_data$temp, main = "Histogram of Temperature", xlab = "Temperature")

# Bar plot of month
barplot(table(forest_fire_data$month), main = "Bar Plot of Month", xlab = "Month", ylab = "Frequency", col = "skyblue")

# Multivariate Plots
# Scatter plot of temperature vs. humidity
plot(forest_fire_data$temp, forest_fire_data$RH, main = "Scatter Plot of Temperature vs. Humidity", xlab = "Temperature", ylab = "Humidity", col = "blue")

# Box plot of temperature across different months
boxplot(temp ~ month, data = forest_fire_data, main = "Box Plot of Temperature Across Months", xlab = "Month", ylab = "Temperature", col = "lightgreen")
```

# Preprocessing & Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values
missing_values <- sum(is.na(forest_fire_data))

# Display summary of missing values
print(paste("Number of missing values:", missing_values))

# Summary of missing values by column
print(summary(is.na(forest_fire_data)))
```

## Transformation
```{r transformation}
# Transform "area" variable using ln(x + 1) function
forest_fire_data$ln_area <- log(forest_fire_data$area + 1)

# Display the structure of the dataset to verify the transformation
str(forest_fire_data)
```

# Training Model
## Data Splitting
```{r Data Splitting}
library(caret)

# Data Splitting
set.seed(123) # Set seed for reproducibility
train_index <- sample(1:nrow(forest_fire_data), 0.8 * nrow(forest_fire_data)) # 80% for training
train_data <- forest_fire_data[train_index, ]
test_data <- forest_fire_data[-train_index, ]

# Print the dimensions of the training and testing sets
print("Dimensions of Training Data:")
print(dim(train_data))
print("Dimensions of Testing Data:")
print(dim(test_data))

```

## Bootstrapping
```{r Bootstrapping}
# Bootstrapping
bootstrap_samples <- lapply(1:1000, function(i) {
  bootstrap_index <- sample(1:nrow(forest_fire_data), replace = TRUE)
  bootstrap_data <- forest_fire_data[bootstrap_index, ]
})

# Calculate mean and median for each bootstrap sample
bootstrap_stats <- sapply(bootstrap_samples, function(sample_data) {
  mean_value <- mean(sample_data$area)
  median_value <- median(sample_data$area)
  return(c(mean_value, median_value))
})

# Calculate confidence intervals for mean and median
confidence_intervals <- t(sapply(bootstrap_stats, function(stat) {
  quantile(stat, c(0.025, 0.975))
}))

# Print confidence intervals
print("Confidence intervals for mean and median:")
print(confidence_intervals)
```

## Cross-validation
```{r Cross-validation}
# Remove the "month" column from the dataset
forest_fire_data <- forest_fire_data[, !(names(forest_fire_data) %in% c("month", "day"))]

# Cross-validation with evaluation using MAD and RMSE after removing the "month" column
library(caret)
set.seed(123) # Set seed for reproducibility
folds <- createFolds(forest_fire_data$area, k = 10) # 10-fold cross-validation

# Placeholder for storing evaluation results
evaluation_results <- list()

for (fold in 1:length(folds)) {
  train_index <- unlist(folds[-fold])
  test_index <- unlist(folds[fold])
  train_data <- forest_fire_data[train_index, ]
  test_data <- forest_fire_data[test_index, ]
  
  # Perform model training (replace this with your model training code)
  model <- lm(area ~ ., data = train_data)
  
  # Make predictions
  predictions <- predict(model, newdata = test_data)
  
  # Evaluate model performance
  MAD <- mean(abs(test_data$area - predictions))
  RMSE <- sqrt(mean((test_data$area - predictions)^2))
  
  # Store evaluation results
  evaluation_results[[paste("Fold", fold)]] <- c(MAD = MAD, RMSE = RMSE)
}
```

## Training Different Models
```{r Training Different Models}
# Load required libraries
library(caret)
library(xgboost)

# Define training control
ctrl <- trainControl(method = "cv",  # Use cross-validation
                     number = 10)    # Number of folds

# Define models
models <- c("lm", "glm", "xgbLinear")

# Train and evaluate each model
results <- lapply(models, function(model) {
  set.seed(123)  # Set seed for reproducibility
  
  # Define formula for regression
  formula <- as.formula(paste("area ~ ."))
  
  # Train model
  if (model == "xgbLinear") {
    # For xgbLinear, convert data to DMatrix format
    dtrain <- xgb.DMatrix(data = as.matrix(forest_fire_data[, -which(names(forest_fire_data) == "area")]), label = forest_fire_data$area)
    fit <- xgboost(data = dtrain, objective = "reg:linear", booster = "gblinear", nrounds = 10, verbose = FALSE)
  } else {
    fit <- train(formula, 
                 data = forest_fire_data, 
                 method = model,
                 trControl = ctrl)
  }
  
  # Make predictions
  if (model == "xgbLinear") {
    predictions <- predict(fit, newdata = as.matrix(forest_fire_data[, -which(names(forest_fire_data) == "area")]))
  } else {
    predictions <- predict(fit, newdata = forest_fire_data)
  }
  
  # Compute evaluation metrics (e.g., RMSE, MAD)
  accuracy <- sqrt(mean((forest_fire_data$area - predictions)^2))
  mad <- mean(abs(forest_fire_data$area - predictions))
  
  # Return model and evaluation metrics
  return(list(model = model, RMSE = accuracy, MAD = mad))
})

# Print results
print("Model performance:")
print(do.call(rbind, results))


# Print evaluation results
print("Evaluation results for each fold:")
print(do.call(rbind, evaluation_results))

```

## Saving Model
```{r Saving Model}
# Load the saved xgbLinear model
loaded_forest_fire_xgb_linear_model <- readRDS("./models/forest_fire_xgb_linear_model.rds")

# Prepare new data for prediction (replace with your actual new data)
new_forest_fire_data <- data.frame(
  X = c(7, 7, 7),  # Example X values
  Y = c(5, 4, 4),  # Example Y values
  month = c("mar", "oct", "oct"),  # Example month values
  day = c("fri", "tue", "sat"),  # Example day values
  FFMC = c(86.2, 90.6, 90.6),  # Example FFMC values
  DMC = c(26.2, 35.4, 43.7),  # Example DMC values
  DC = c(94.3, 669.1, 686.9),  # Example DC values
  ISI = c(5.1, 6.7, 6.7),  # Example ISI values
  temp = c(8.2, 18, 14.6),  # Example temp values
  RH = c(51, 33, 33),  # Example RH values
  wind = c(6.7, 0.9, 1.3),  # Example wind values
  rain = c(0, 0, 0)  # Example rain values
)

# Convert new data to matrix
new_data_matrix <- as.matrix(new_forest_fire_data[, -which(names(new_forest_fire_data) %in% c("month", "day"))])

# Use the loaded model to make predictions for new forest fire data
predictions_xgb_loaded_model <- predict(loaded_forest_fire_xgb_linear_model, newdata = xgb.DMatrix(new_data_matrix))


# Apply inverse of ln(x+1) transform
predictions_xgb_loaded_model <- exp(predictions_xgb_loaded_model) - 1

# Print predictions
print(predictions_xgb_loaded_model)

```


